---
title: "*Vanessa cardui* African divide"
output: html_notebook
---
```{r}
library(topGO)
```

## topGo analysis for inverted region

### Preparing data

Preparing functional annotation:
```{r}
fun_ann <- read.csv(file ="/Users/dshipilina/GitHub/VanessaExpression/vcard.func.annot.csv",header = FALSE)
names(fun_ann)[names(fun_ann)=="V1"]<-"GeneID"
names(fun_ann)[names(fun_ann)=="V8"]<-"Function"
names(fun_ann)[names(fun_ann)=="V9"]<-"GeneName"
names(fun_ann)[names(fun_ann)=="V21"]<-"Func_Categ"
fun_ann$GeneID<-gsub("-RA","",as.character(fun_ann$GeneID))
functional_annotation<-fun_ann[,c("GeneID","Function","GeneName","Func_Categ")]

#functional_annotation[is.element(functional_annotation$GeneID,c(special_genes)),]

```

Preparing genes of interest
```{bash}
#awk '$1 == "LR999932.1" && $3 == "mRNA" && $4 >= 4100000 && $5 <= 13100000 {print $9} ' makerrun3.all.maker.rename.proteins.AED50.eAED50.long50.norepeatdomain.no1exon.nooverlap.noW.gff | awk -F ";" '{print $1}'| sed 's/ID=//' > inversion_genes.tsv

grep -f inversion_genes.tsv /Users/dshipilina/GitHub/VanessaExpression/vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}' > inversion_genes_func.tsv
```

```{r}
topGOinversion<-Dasha_topGO('inversion_genes.tsv')
head(topGOinversion)
KarinsMagic_topGO(topGOinversion)
ggsave(plot = KarinsMagic_topGO(topGOinversion), "topGOinversion.pdf")
```
```{r}

topGOinversion$GO_descr <- paste(topGOinversion$Term," (",topGOinversion$GO.ID,")", sep = "")
plot_topGo=ggplot(topGOinversion, aes(reorder(GO_descr, -Significant), Significant)) +
    geom_bar(stat = "identity", aes(fill=GO_class))  +
  ylab("Number of genes (p-value < 0.01)") +   facet_grid(~GO_class, scales = "free_x", space = "free_x") +
  scale_fill_discrete(guide = guide_legend(label.position = "top")) +
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 3, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 3, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 3, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 3, angle = 90, vjust = 0.5),
        legend.position = "top")

ggsave(plot = plot_topGo, "V_cardui_inversion.pdf")
plot_topGo
```
```{r}
topGOinversion
topGOinversionBP<-topGOinversion[(topGOinversion$GO_class == "BP"),]

topGOinversionBP$GO_descr <- paste(topGOinversionBP$Term," (",topGOinversionBP$GO.ID,")", sep = "")

plot_topGo=ggplot(topGOinversionBP, aes(x=reorder(GO_descr, -Significant), y=Significant)) + geom_bar(stat="identity") + 
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 3, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 3, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 3, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 3, angle = 90, vjust = 0.5),
        legend.position = "top",
        axis.ticks = element_line(colour = "black", size = 0.1))

ggsave(plot = plot_topGo, "V_cardui_inversion.pdf")
plot_topGo
```
```
+
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 6, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 6, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 6, angle = 90, vjust = 0.5),
        legend.position = "top")


```{r}
#Manually selected genes

special_genes<-c("Vcard_DToL07128","Vcard_DToL04008","Vcard_DToL02159","Vcard_DToL10306","Vcard_DToL05734","Vcard_DToL07345","Vcard_DToL07927","Vcard_DToL12485")

par(mfrow=c(2,3))
for (x in special_genes) {
  plotCounts(ddsE1, gene=x, intgroup="condition")
}

functional_annotation[is.element(functional_annotation$GeneID,c(special_genes)),]
```

```{r}
#read the list of genes of interest res05hef2topDowngenenames
resE105topUp <- resE1[resE1$padj <= 0.05,]
y<-row.names(resE105topUp)
write.csv(y,file="E1genenames.tsv",row.names=F)
```

```{bash}
tail -n +2 E1genenames.tsv | sed 's/\"//' | sed 's/\"//' | awk '{print $1"-RA"}' > E1topGogenenames.tsv
```

```{r}
allRes_E1<-Dasha_topGO('E1topGogenenames.tsv')
head(allRes_E1)
KarinsMagic_topGO(allRes_E1)
ggsave(plot = KarinsMagic_topGO(allRes_E1), "V_cardui_E1.pdf")

```





#### Contrast 2: HDAL vs HDLI, Pupa, Heads, Sexes as factor, Family as Factor


```{r}
resE1105<-results(ddsE1, contrast=c("condition","HDAL","HDLI"),alpha=0.05)
resE11<-results(ddsE1, contrast=c("condition","HDAL","HDLI"))
summary(resE11)
resE1105 <- resE1105[order(resE1105$padj),]
head(resE1105)

#Different shrinkage
#resE1LFC <- lfcShrink(ddsE1, coef="condition_LDAL_vs_HDAL", type="normal")
#resE1LFCa <- lfcShrink(ddsE1, coef="condition_LDAL_vs_HDAL", type="apeglm")
#resLFC

plotMA(resE11, ylim=c(-20,20))
#plotMA(resE1LFC, ylim=c(-3,3))
#plotMA(resE1LFCa, ylim=c(-2,2))
#plotMA(res05abf)#, ylim=c(-2,2))
```
```{r}
# Make a basic volcano plot
with(resE1105, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot", xlim=c(-4,4)))

# Add colored points: blue if padj<0.05, red if log2FC>1 and padj<0.05)
with(subset(resE1105, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(resE1105, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))
```


```{r}
resE1105<-na.omit(resE1105)
resE1105topUp <- resE1105[resE1105$padj <= 0.05 & resE1105$log2FoldChange > 0, ]
resE1105topDown <- resE1105[resE1105$padj <= 0.05 & resE1105$log2FoldChange < 0, ]
head(resE1105topUp,nrow(resE1105topUp))
head(resE1105topDown,nrow(resE1105topDown))


par(mfrow=c(1,2))
for (x in row.names(resE1105topUp)) {
  plotCounts(ddsE1, gene=x, intgroup="condition")
}

par(mfrow=c(1,2))
for (x in row.names(resE1105topDown)) {
  plotCounts(ddsE1, gene=x, intgroup="condition")
}

par(mfrow=c(1,2))
for (x in row.names(resE1105topUp)) {
  plotCounts(ddsE1, gene=x, intgroup="sex")
}

par(mfrow=c(1,2))
for (x in row.names(resE1105topDown)) {
  plotCounts(ddsE1, gene=x, intgroup="sex")
}

par(mfrow=c(1,2))
for (x in row.names(resE1105topDown)) {
  plotCounts(ddsE1, gene=x, intgroup="family")
}

par(mfrow=c(1,2))
for (x in row.names(resE1105topUp)) {
  plotCounts(ddsE1, gene=x, intgroup="family")
}

```


```{r}
#Manually selected genes

special_genes_ALLI<-c("Vcard_DToL15090","Vcard_DToL03953","Vcard_DToL07345","Vcard_DToL15312","Vcard_DToL05602","Vcard_DToL07835","Vcard_DToL02158","Vcard_DToL04968","Vcard_DToL07353","Vcard_DToL09031")

par(mfrow=c(2,3))
for (x in special_genes_ALLI) {
  plotCounts(ddsE1, gene=x, intgroup="condition")
}

functional_annotation[is.element(functional_annotation$GeneID,c(special_genes_ALLI)),]
```


```{r}
#read the list of genes of interest res05hef2topDowngenenames
resE1105topUp <- resE1105[resE1105$padj <= 0.05,]
y<-row.names(resE1105topUp)
write.csv(y,file="E11genenames.tsv",row.names=F)
```

```{bash}
tail -n +2 E11genenames.tsv | sed 's/\"//' | sed 's/\"//' | awk '{print $1"-RA"}' > E11topGogenenames.tsv
```





```{r}
#read the list of genes of interest 
#"when only a list of interesting genes is provided, 
#the user can use only tests statistics that are based on gene counts, 
#like Fisher’s exact test, Z score and alike."
##Algorithms:
#classic uses all GO terms separately
#elim eliminates genes from parentGOterms if child is more specific (bottom up analysis)
#weight trying to determin if GOterm is better representing the list of interesting genes (is more enriched) than any other node from its neighborhood
#parentChild 

Dasha_topGO <- function(gene_list){

candidate_genes <- read.table(gene_list, header = F)
colnames(candidate_genes) <- "geneid"
#head(candidate_genes)

candidate_genes <- as.character(candidate_genes$geneid)
geneList <- factor(as.integer(geneNames %in% candidate_genes))
names(geneList) <- geneNames
#head(geneList)

##Biological function
#create the topGOdata object
GO_data_BP <- new("topGOdata", 
               ontology="BP", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)

BP_resultFisher_weight01 <- runTest(GO_data_BP, algorithm = "weight01", statistic = "fisher")

allGO = usedGO(object = GO_data_BP)
allRes_BP <- GenTable(GO_data_BP, 
                   weight01Fisher = BP_resultFisher_weight01,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

#multiple test correction, method Benjamini-Hochberg
allRes_BP$p_adj <- p.adjust(allRes_BP$weight01Fisher, method = "fdr")
allRes_BP$GO_class <- "BP"

###Molecular function
#create the topGOdata object
GO_data_MF <- new("topGOdata", 
               ontology="MF", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)

MF_resultFisher_weight01 <- runTest(GO_data_MF, algorithm = "weight01", statistic = "fisher")

allGO = usedGO(object = GO_data_MF)
allRes_MF <- GenTable(GO_data_MF, 
                   weight01Fisher = MF_resultFisher_weight01,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_MF$weight01Fisher <- as.numeric(allRes_MF$weight01Fisher)

#multiple test correction, method Benjamini-Hochberg
allRes_MF$p_adj <- p.adjust(allRes_MF$weight01Fisher, method = "fdr")
allRes_MF$GO_class <- "MF"

###Cellular compartment
#create the topGOdata object
GO_data_CC <- new("topGOdata", 
               ontology="CC", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)

CC_resultFisher_weight01 <- runTest(GO_data_CC, algorithm = "weight01", statistic = "fisher")

allGO = usedGO(object = GO_data_CC)
allRes_CC <- GenTable(GO_data_CC, 
                   weight01Fisher = CC_resultFisher_weight01,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_CC$weight01Fisher <- as.numeric(allRes_CC$weight01Fisher)
#multiple test correction, method Benjamini-Hochberg
allRes_CC$p_adj <- p.adjust(allRes_CC$weight01Fisher, method = "fdr")
allRes_CC$GO_class <- "CC"

#assembling all the data
allRes_all_comb <- rbind(allRes_BP[allRes_BP$weight01Fisher<=0.05, ], allRes_CC[allRes_CC$weight01Fisher<=0.05, ], allRes_MF[allRes_MF$weight01Fisher<=0.05, ])

return(allRes_all_comb)

}
```

```{r}
allRes_all_comb<-Dasha_topGO('E11topGogenenames.tsv')
```


```{r}
#allRes_all_comb <- rbind(allRes_BP[allRes_BP$weight01Fisher<=0.05, ], allRes_CC[allRes_CC$weight01Fisher<=0.05, ], allRes_MF[allRes_MF$weight01Fisher<=0.05, ])

KarinsMagic_topGO <- function(allRes_all_comb){
  
allRes_all_comb$GO_descr <- paste(allRes_all_comb$Term," (",allRes_all_comb$GO.ID,")", sep = "")

#plot go-terma and number of significant genes
plot_topGo=ggplot(allRes_all_comb, aes(reorder(GO_descr, -Significant), Significant)) +
    geom_bar(stat = "identity", aes(fill=GO_class)) +
  facet_grid(~GO_class, scales = "free_x", space = "free_x") +
  ylab("Number of genes (p-value < 0.01)") +
  scale_fill_discrete(guide = guide_legend(label.position = "top")) +
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 5, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 5, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 5, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 5, angle = 90, vjust = 0.5),
        legend.position = "top"
        )

#ggsave(plot = plot_topGo, "V_cardui_E11.pdf")
return(plot_topGo)

}
```



```{r}
KarinsMagic_topGO(allRes_all_comb)
```


```{r}
#allRes_all_comb <- rbind(allRes_BP[allRes_BP$weight01Fisher<=0.05, ], allRes_CC[allRes_CC$weight01Fisher<=0.05, ], allRes_MF[allRes_MF$weight01Fisher<=0.05, ])

allRes_all_comb$GO_descr <- paste(allRes_all_comb$Term," (",allRes_all_comb$GO.ID,")", sep = "")

#plot go-terma and number of significant genes
plot_topGo=ggplot(allRes_all_comb, aes(reorder(GO_descr, -Significant), Significant)) +
    geom_bar(stat = "identity", aes(fill=GO_class)) +
  facet_grid(~GO_class, scales = "free_x", space = "free_x") +
  ylab("Number of genes (p-value < 0.01)") +
  scale_fill_discrete(guide = guide_legend(label.position = "top")) +
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 5, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 5, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 5, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 5, angle = 90, vjust = 0.5),
        legend.position = "top"
        )

ggsave(plot = plot_topGo, "V_cardui_E11.pdf")
plot_topGo
```


#### Contrast 3: HDAL vs LDAL, Adult, Heads, Sexes as factor, Family as Factor

```{r}
colDataE3<-colDataHDLD[(colDataHDLD$stage == "ADM" | colDataHDLD$stage == "ADF" | colDataHDLD$stage == "ADMA"),] #selecting pupa samples only
#head(colDataE1)
colDataE3<-colDataE3[colDataE3$tissue == "H", ] #selecting head samples only
#head(colDataE1)


##Adding sex information
colDataE3$sex <- colDataE3$stage
colDataE3$sex<-gsub("AD","",as.character(colDataE3$sex))
colDataE3$sex<-gsub("A","",as.character(colDataE3$sex))
colDataE3$stage<-gsub("F","",as.character(colDataE3$stage))
colDataE3$stage<-gsub("MA","",as.character(colDataE3$stage))
colDataE3$stage<-gsub("M","",as.character(colDataE3$stage))

##Selecting count data
sample_names<-colDataE3$sample
countE3<-countHDLD[,c(sample_names)]

head(colDataE3)
head(countE3)
```


```{r}
ddsE3 <- DESeqDataSetFromMatrix(countData=countE3, colData =colDataE3, design = ~family + sex + condition)

#Basic filtering
keep <- rowSums(counts(ddsE3)) >= 10
ddsE3 <- ddsE3[keep,]
head(ddsE3)

```



```{r}
ddsE3 <- DESeqDataSetFromMatrix(countData=countE3, colData =colDataE3, design = ~family + sex + condition)

#Basic filtering
keep3 <- rowSums(counts(ddsE3)) >= 10
ddsE3 <- ddsE3[keep3,]
head(ddsE3)

#Running DeSeQ
ddsE3 <- DESeq(ddsE3)

#Summary of results function
#Setting up p-value to 0.05
resE3 <- results(ddsE3)
resE305 <- results(ddsE3, alpha=0.05)
summary(resE305)
resE305 <- resE305[order(resE305$padj),]
head(resE305)

#Different shrinkage
resE3LFC <- lfcShrink(ddsE3, coef="condition_LDAL_vs_HDAL", type="normal")
resE3LFCa <- lfcShrink(ddsE3, coef="condition_LDAL_vs_HDAL", type="apeglm")
#resLFC

plotMA(resE3, ylim=c(-20,20))
plotMA(resE3LFC, ylim=c(-3,3))
plotMA(resE3LFCa, ylim=c(-2,2))
#plotMA(res05abf)#, ylim=c(-2,2))

# Make a basic volcano plot
with(resE3, plot(log2FoldChange, -log10(pvalue), pch=20, main="condition_LDAL_vs_HDAL_pupa_head", xlim=c(-4,4),ylim=c(0,6)))

# Add colored points: blue if padj<0.05, red if log2FC>1 and padj<0.05)
with(subset(resE3, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(resE3, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))
```


```{r}
ddsE3 <- DESeqDataSetFromMatrix(countData=countE3, colData =colDataE3, design = ~family + sex + condition)

#Basic filtering
keep3 <- rowSums(counts(ddsE3)) >= 10
ddsE3 <- ddsE3[keep3,]
head(ddsE3)

#Running DeSeQ
ddsE3 <- DESeq(ddsE3)

#Summary of results function
#Setting up p-value to 0.05
resE3 <- results(ddsE3)
resE305 <- results(ddsE3, alpha=0.05)
summary(resE305)
resE305 <- resE305[order(resE305$padj),]
head(resE305)

#Different shrinkage
resE3LFC <- lfcShrink(ddsE3, coef="condition_LDAL_vs_HDAL", type="normal")
resE3LFCa <- lfcShrink(ddsE3, coef="condition_LDAL_vs_HDAL", type="apeglm")
#resLFC

plotMA(resE3, ylim=c(-20,20))
plotMA(resE3LFC, ylim=c(-3,3))
plotMA(resE3LFCa, ylim=c(-2,2))
#plotMA(res05abf)#, ylim=c(-2,2))

# Make a basic volcano plot
with(resE3, plot(log2FoldChange, -log10(pvalue), pch=20, main="condition_LDAL_vs_HDAL_pupa_head", xlim=c(-4,4),ylim=c(0,6)))

# Add colored points: blue if padj<0.05, red if log2FC>1 and padj<0.05)
with(subset(resE3, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(resE3, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))
```

```{r}

candidate_genes<-function(dds,res){

res<-na.omit(res)
restopUp <- res[res$padj <= 0.05 & res$log2FoldChange > 0, ]
restopDown <- res[res$padj <= 0.05 & res$log2FoldChange < 0, ]
#head(restopUp,nrow(restopUp))
#head(restopDown,nrow(restopDown))


par(mfrow=c(1,2))
for (x in row.names(restopUp)) {
  plotCounts(dds, gene=x, intgroup="condition")
}

par(mfrow=c(1,2))
for (x in row.names(restopDown)) {
  plotCounts(dds, gene=x, intgroup="condition")
}

par(mfrow=c(1,2))
for (x in row.names(restopUp)) {
  plotCounts(dds, gene=x, intgroup="sex")
}

par(mfrow=c(1,2))
for (x in row.names(restopDown)) {
  plotCounts(dds, gene=x, intgroup="sex")
}

par(mfrow=c(1,2))
for (x in row.names(restopDown)) {
  plotCounts(dds, gene=x, intgroup="family")
}

par(mfrow=c(1,2))
for (x in row.names(restopUp)) {
  plotCounts(dds, gene=x, intgroup="family")
}

}

```


```{r}
candidate_genes(ddsE3,resE3)
```





